---
title: "Setting up credential-handling in trex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setting up credential-handling in trex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(trex)
```

Some APIs supported by trex will accept credentials identifying the level of access that the user has been granted which can be used to request non-public data from the APIs. Passwords, API keys, and API tokens are all stored by trex in a system-managed encrypted keyring (using the package [keyring](https://cran.r-project.org/web/packages/keyring/index.html)) to avoid exposing sensitive data.

If you do not have special permissions for data access through any APIs, this document is not relevant and can be ignored.

# Initial Setup

The first step is to set up a keyring for trex to use. If the system supports multiple keyrings, this will be specific to the username you provide and trex but otherwise trex will use the default system keyring. If you have not set up a keyring, you will receive a console message on loading the package that suggests to creating one.

``` R
# Run the function setup_keyring() with the username associated with the API.
setup_keyring(username = "smokey.bear@usda.gov")
```

When creating a keyring, you will be prompted to enter a password. This is the password that will be used to unlock the keyring *not* the password associated with the API. It's best practice to keep those separate. When running trex functions that look for API credentials, this is the password that you will need to provide to unlock the keyring and retrieve whichever credentials are needed.

# Storing Credentials

There are three kinds of credentials that can be stored and referenced by trex:

-   Passwords associated with API accounts

-   API keys

-   API tokens

## Passwords

To store a password, you need to run `store_password()` providing the keyring username and the name of the API. The function does not take the password as an argument and instead will prompt you to interactively enter the password which is then stored in the keyring associated with the username.

``` R
# Store a password for the LDC API account associated with smokey.bear@usda.gov
store_password(username = "smokey.bear@usda.gov",
               api_name = "ldc")

# In cases where there is already a stored password for that API account, you can
# replace it by using overwrite = TRUE
store_password(username = "smokey.bear@usda.gov",
               api_name = "ldc",
               overwrite = TRUE)
```

## API Keys

Some APIs have users generate keys which have specific permissions and limited lifespans, typically days to months. These are static alphanumeric strings that can be submitted with a request to an API and are often created and managed in a web interface associated with the API.

To store an API key, you need to run `store_api_key()` providing the username and the name of the API key. It's important to choose an api_key_name value that you can remember easily because you may have multiple API keys with different permissions and if you don't specify a name, the key will be called "default". Best practice is to give a key the same name here that it has in the API interface if allowed.

The function does not take the API key as an argument and instead will prompt you to interactively enter the key which is then stored in the keyring associated with the username. API keys tend to be long, so you should copy and paste the key instead of re-typing it if possible but make sure to not include any characters that aren't part of the key, e.g., quotation marks or spaces at the ends.

``` R
# Store a key for the LDC API account associated with smokey.bear@usda.gov.
# This key is being named "basic_access" because it has no special permissions.
store_api_key(username = "smokey.bear@usda.gov",
              api_key_name = "basic_access")

# Store another key for the LDC API account associated with smokey.bear@usda.gov.
# This key is being named "unrestricted" because it has full permissions.
store_api_key(username = "smokey.bear@usda.gov",
              api_key_name = "unrestricted")

# In the case where there is already a stored API key with that name for that API
# account, you can replace it by using overwrite = TRUE.
store_password(username = "smokey.bear@usda.gov",
               api_key_name = "unrestricted",
               overwrite = TRUE)
```

## API Tokens

API tokens are another way that APIs manage limited permissions with expiration times, often measured in hours. Because they're so short-lived and a new token can be requested from a server using your username and stored password, you're unlikely to be storing them directly. That said, the functions that trex uses internally for token management are also available to end users.

If you need to store a new token directly, you run `store_api_token()` , providing the API username and the name of the API. The function will prompt you for your API password which it then uses with the username to request a token and store it like other credentials.

``` R
# This will request a token from the LDC API for the account associated with
# smokey.bear@usda.gov and then store it.
store_api_token(username = "smokey.bear@usda.gov",
                api_name = "ldc")

# In the case that there's already a stored, unexpired token, you can force requsting
# and storing a new one with overwrite = TRUE
store_api_token(username = "smokey.bear@usda.gov",
                api_name = "ldc",
                overwrite = TRUE)
```

# Using Credentials

In most cases, using credentials only requires the username used to store them, although using an API key also needs the name that the key was stored under. If your system has locked the keyring, you'll be prompted to unlock it again in order for trex to reference the contents.

In order to retrieve data from the Landscape Data Commons using `fetch_ldc()` with the credentials stored in the previous examples, your code would look like this:

``` R
# Getting all canopy gap data permitted with the LDC API key stored under the name
# "unrestricted"
gap_data <- fetch_ldc(data_type = "gap",
                      username = "smokey.bear@usda.gov",
                      api_key_name = "unrestricted")

# Getting all the soil stability data permitted with a token associated with the
# username smokey.bear@usda.gov.
# Because tokens associated with an account all have the same permissions, all that's
# needed it to set token = TRUE for the function to check for a valid stored token
# and request a new one if needed.
soil_stability_data <- fetch_ldc(data_type = "soilstability",
                                 username = "smokey.bear@usda.gov",
                                 token = TRUE)
```
