---
title: "Using Ecosystem Dynamics Interpretive Tool functions"
author: "Joe Brehm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Ecosystem Dynamics Interpretive Toolfunctions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The Ecosystem Dynamics Interpretive Tool (EDIT) provides detailed data on ecological site descriptions and state and transition models. These are available via [website](https://edit.jornada.nmsu.edu/), and via API. `trex` provides tools to easily access these data, via the function `fetch_edit()`. 

## fetch_edit()

This is the overarching wrapper function for making any query of the EDIT API. User-provided parameters specify the data to return, and the MLRA(s) to return data for.


### Basic use

The EDIT function requires at minimum the `mlra` and `data_type` arguments. All ecological sites within an MLRA will be queried for that MLRA, returning the requested data type. Valid `data_type` inputs are "ecosites", "rangeland", "overstory", "understory". "climatic-features", "ecological-dynamics", "general-information", "interpretations", "physiographic-features", "reference-sheet", "soil-features", "supporting-information", "water-features", and "states". The function contains a `verbose` parameter, if TRUE it will provide diagnostic messages that include the URLs the function is attempting to query. 

```{r, eval=FALSE}
# This retrieves ecological site IDs and names from MLRA 039X, and reports queried URLs.
edit_039X_ecosite <- fetch_edit(mlra = "039X", data_type = "ecosite", verbose = TRUE)

# This retrieves ecological site IDs and names from MLRAs 039X and 040X, and reports queried URLs.
edit_039X040X_ecosite <- fetch_edit(mlra = c("039X", "040X"), data_type = "ecosite", verbose = TRUE)

# This retrieves climate data from MLRA 039X. It does not report queried URLs, as this is a large set of queries. 
edit_039X_climate <- fetch_edit(mlra = "039X", data_type = "climate", verbose = FALSE)

```

### Advanced use

#### keys and key_type
EDIT ecological sites can be queried by a set of request parameters using `keys` and `key_type`, filtering the returned data by properties of the ecological sites. Due to the nature of EDIT, there are a limited range of parameters that can be queried. These are: "precipitation", "frostFreeDays", "elevation", "slope", "landform", "parentMaterialOrigin", "parentMaterialKind", and "surfaceTexture". Note that when querying numeric parameters, the ecological sites returned include all of those that occur within a given range of parameters, for example a query of slope from between 10 and 20% would include a site that occurs between 15 and 30% slope.

All `data_type` options are queriable with the same set of parameters.

```{r, eval = FALSE}
# This retrieves ecological site IDs and names from MLRA 039X, for sites that occur at slopes between 15 and 30%, and reports queried URLs.
edit_039X_ecosite_filter_slope <- fetch_edit(mlra = "039X", data_type = "ecosite", keys = "15:30", key_type = "slope", verbose = TRUE)

# This retrieves ecological site IDs and names from MLRA 039X, for sites that occur at slopes between 15 and 30%, and reports queried URLs.
edit_039X_ecosite_filter_landform <- fetch_edit(mlra = "039X", data_type = "ecosite", keys = "mountain", key_type = "landform", verbose = TRUE)

```

#### key_chunk_size

This argument determines how many of the provided keys will be used for a single API query. There's a character count limit for queries and so it's possible that a requesting data via a single query could exceed that limit. By setting `key_chunk_size`, you can control how many keys will be used in a query at a time. The default is `100` but you may find that you can raise that or need to lower it. Regardless, all the keys you provide will be used in a query and the data frame returned by the function will contain the results from every query. If you use `verbose = TRUE` you can see each query being made as the function reports back.

Unlike with the LDC, EDIT queries will rarely require this parameter to be changed, but it is an option available for troubleshooting.

```{r, eval=FALSE}
# This retrieves overstory species community data using a smaller chunk size.
edit_039X_overstory_50 <- fetch_edit(mlra = "039X",
                                     data_type = "overstory",
                                     key_chunk_size = 50,
                                     verbose = FALSE)

```

#### timeout

It's possible that the server may be slow to respond to a request, in which case you can adjust the time that the function waits before considering a nonresponse to be a failure. This is defined in seconds via the argument `timeout` and defaults to `60`. This is not likely to ever need to be changed because 60 seconds is very generous, but is available as an option.
