---
title: "Using Ecosystem Dynamics Interpretive Tool functions"
author: "Joe Brehm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Ecosystem Dynamics Interpretive Tool functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The Ecosystem Dynamics Interpretive Tool (EDIT) provides detailed data on ecological site descriptions and state and transition models. These are available via [website](https://edit.jornada.nmsu.edu/), and via API. `trex` provides tools to easily access these data, via the function `fetch_edit()`. 

## fetch_edit()

This is the function for making any query of the EDIT API. User-provided parameters specify the data to return, and the MLRA(s) to return data for.

### Basic use

The EDIT function requires at minimum the `mlra` and `data_type` arguments. All ecological sites within an MLRA will be queried for that MLRA, returning the requested data type. Valid `data_type` inputs are "ecosites", "climate", "landforms", "physiography interval", "physiography nominal", "physiography ordinal", "annual production", "overstory", "understory", "rangeland", "surface cover", "parent material", "soil interval", "soil nominal", "soil ordinal", "soil profile", "texture", "state narratives", "transition narratives", and "mlra"*. The function contains a `verbose` parameter, if TRUE it will provide diagnostic messages that include the URLs the function is attempting to query. 

* data_type "mlra" is a special case, returning all MLRA codes present in EDIT. This is intended as a way to look up the precise codes -- e.g. MLRA 42 cannot be queried, but MLRA 042X can be. To use this functionality, set argument `mlra` to NULL. Or, more simply, use the convenience function fetch_mlra_codes() with no arguments.

For information about what these tables contain, see the EDIT developer resources page: https://edit.jornada.nmsu.edu/resources/esd, under the tab-delimited text section. 

```{r, eval=FALSE}
# This retrieves ecological site IDs and names from MLRA 039X, and reports queried URLs.
edit_039X_ecosite <- fetch_edit(mlra = "039X", data_type = "ecosites", verbose = TRUE)

# This retrieves ecological site IDs and names from MLRAs 039X and 040X, and reports queried URLs.
edit_039X040X_ecosite <- fetch_edit(mlra = c("039X", "040X"), data_type = "ecosites", verbose = TRUE)

# This retrieves climate data from MLRA 039X, and reports queried URLs.
edit_039X_climate <- fetch_edit(mlra = "039X", data_type = "climate", verbose = TRUE)

```

### Advanced use

#### Finding MLRA codes
To use fetch_edit(), you must know the exact MLRA code. To find a list of exact codes used in EDIT, use the function fetch_mlra_codes(). This can also be done with fetch_edit() by setting mlra to `NULL` and data_type to `mlra`, but fetch_mlra_codes() provides an intuitive wrapper.

```{r, eval = FALSE}
fetch_mlra_codes()
```

#### keys and key_type
EDIT ecological sites can be queried by a set of request parameters using `keys` and `key_type`, filtering the returned data by properties of the ecological sites. Due to the nature of EDIT, there are a limited range of parameters that can be queried. These are: "precipitation", "frostFreeDays", "elevation", "slope", "landform", "parentMaterialOrigin", "parentMaterialKind", and "surfaceTexture". Note that when querying numeric parameters, the ecological sites returned include all of those that occur within a given range of parameters, for example a query of slope from between 10 and 20% would include a site that occurs between 15 and 30% slope.

All `data_type` options are queriable with the same set of parameters, except for `mlra`. 

```{r, eval = FALSE}
# This retrieves ecological site IDs and names from MLRA 039X, for sites that occur at slopes between 15 and 30%, and reports queried URLs.
edit_039X_ecosite_filter_slope <- fetch_edit(mlra = "039X", data_type = "ecosites", keys = "15:30", key_type = "slope", verbose = TRUE)

# This retrieves ecological site IDs and names from MLRA 039X, for sites that occur at slopes between 15 and 30%, and reports queried URLs.
edit_039X_ecosite_filter_landform <- fetch_edit(mlra = "039X", data_type = "ecosites", keys = "mountain", key_type = "landform", verbose = TRUE)

```

#### tall vs wide output
Data can be returned in either a tall/long format, or a wide format. To control this, use the argument `tall`. This defaults to `TRUE`, returning tall data.

#### key_chunk_size

This argument determines how many of the provided keys will be used for a single API query. There's a character count limit for queries and so it's possible that a requesting data via a single query could exceed that limit. By setting `key_chunk_size`, you can control how many keys will be used in a query at a time. The default is `100` but you may find that you can raise that or need to lower it. Regardless, all the keys you provide will be used in a query and the data frame returned by the function will contain the results from every query. If you use `verbose = TRUE` you can see each query being made as the function reports back.

Unlike with the Landscape Data Commons, EDIT queries will rarely require this parameter to be changed, but it is an option available for troubleshooting.

```{r, eval=FALSE}
# This retrieves overstory species community data using a smaller chunk size.
edit_003X_overstory_50 <- fetch_edit(mlra = "003X",
                                     data_type = "overstory",
                                     key_chunk_size = 50,
                                     verbose = FALSE)

```

#### timeout

It is possible that the server may be slow to respond to a request, in which case you can adjust the time that the function waits before considering a nonresponse to be a failure. This is defined in seconds via the argument `timeout` and defaults to `60`. This is not likely to ever need to be changed because 60 seconds is very generous, but is available as an option.

#### delay

Very large requests with multiple MLRAs may overload the API server. In order to prevent this, the function will insert a delay between queries. This is defined in milliseconds, and defaults to 500ms. Like `timeout` and `key_chunk_size`, this is unlikely to ever need to be changed.

#### path_unparsable_data

Unfortunately, due to data entry errors, tabs may be present in narrative data. Because the data returned by the EDIT API is tab-delimited, this renders narrative data from certain ecological sites unparsable. When this occurs, the function will output a warning. Data from the offending ecological sites will be excluded from the final output, and will instead be saved to a text file in a folder provided by the argument `path_unparsable_data`. These data can be manually processed, but unfortunately they require a human eye to interpret. To recover these data, edit the text file, removing unnecessary tabs from the text columns, and read the file into R using read.delim. 