---
title: "Using Landscape Data Commons functions"
author: "Nelson Stauffer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Landscape Data Commons functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
```

The Landscape Data Commons (LDC) is a repository of publicly-available ecological data sets. These are accessible via API and `trex` contains functions to help download the data without needing to write API calls yourself. The core function is `fetch_ldc()` and there are two additional functions for common use-cases, `fetch_ldc_spatial()` and `fetch_ldc_ecosite()`.

## fetch_ldc()

This is the main function for accessing the LDC API and is used by both `fetch_ldc_spatial()` and `fetch_ldc_ecosite()`. It takes the user-provided parameters to request data via the API to automatically build the relevant API call(s) and returns the results of the call(s). By default, the function will silently create however many API calls it needs to to retrieve the requested data, but you can always use the argument `verbose = TRUE` to see what steps it's taking.

### Basic use

The minimum information the function requires to work is the `data_type` argument which corresponds to which table in the LDC the data are being requested from. If you specify a data type and nothing else, then the function will attempt to retrieve all data associated with the requested data type. This is not recommended for most data types because the number of records in tables of raw data can be very high. However, it would not be inappropriate to use with `data_type = "header"` or `data_type = "indicators"` because these two tables contain "only" one entry for each sampling event, being the metadata and calculated ecological indicators for each sampling event, respectively.

```{r, eval=FALSE}
# This retrieves all header information from the LDC
# Here verbose = TRUE so that the function will report back on what it's doing
# at each step
all_ldc_headers <- fetch_ldc(data_type = "header",
                             verbose = TRUE)

# To retrieve all calculated ecological indicator values from the LDC
all_ldc_indicators <- fetch_ldc(data_type = "indicators")
```

When retrieving raw data, as from line-point intercept or canopy gap intercept, it is recommended that you also use the argument `query_parameters`. Using this lets you specify variables/columns in the table called for by `data_type` or other tables available through the API. Every table in the LDC has different variables available (although some, like PrimaryKey which connects them all, are found in more than one), so consult [the documentation](https://api.landscapedatacommons.org/docs) if you're uncertain which variables are available.

The "easiest" variable to use is PrimaryKey because it's the one that ties all records from a sampling event together both within and across tables. That does require knowing what PrimaryKey values to use for the query parameters though, so consider using `fetch_ldc_spatial()` or `fetch_ldc_ecosite()` (see sections below) to retrieve records without having to know any PrimaryKey values.

```{r, eval=FALSE}
# To retrieve all line-point intercept records associated with the sampling
# event with the PrimaryKey value of 16060107170093802016-09-01
relevant_pk <- c("16060107170093802016-09-01")
singlepk_query_parameters <- list("PrimaryKey" = list("=" = relevant_pk))
lpi_data_single <- fetch_ldc(data_type = "lpi",
                             query_parameters = singlepk_query_parameters)

# To retrieve all line-point intercept records associated with the sampling
# event with the PrimaryKey value of 20198113201104B2, 2019885200311B1, and
# 2019885200311B2
relevant_pks <- c("20198113201104B2",
                  "2019885200311B1",
                  "2019885200311B2")
multipk_query_parameters <- list("PrimaryKey" = list("=" = relevant_pks))
lpi_data_multiple <- fetch_ldc(data_type = "lpi",
                               query_parameters = multipk_query_parameters)
```

You can query using multiple parameters at a time and with variables that occur in different tables. This can be risky because not every PrimaryKey value occurs in every table, so attempting to query by a variable in a table that does not contain any PrimaryKeys that are turned up in other tables will return no data. For that reason, be very careful if your query parameters use data tables associated with multiple data types that aren't guaranteed to have been collected at all plots you would want results from.

If you find that `fetch_ldc()` is returning errors or unexpected results, try using `verbose = TRUE` which will force the function to report what it's doing at each step, including the full queries that it's sending to the LDC server. This can be indispensible for troubleshooting.

### Advanced use

There are additional arguments that you can supply to `fetch_ldc()` to help, especially if your queries are returning errors from the server.

#### timeout

It's possible that the server may be slow to respond to a request, in which case you can adjust the time that the function waits before considering a nonresponse to be a failure. This is defined in seconds via the argument `timeout` and defaults to `300` which is five minutes. This is not likely to ever need to be changed because five minutes is very generous, but is available as an option.

#### take

In addition to there being a character limit to query length, there is also an effective limit to the number of records that can be returned by a single query. Although there are several ways to get a 500 error from the server, the most likely is that a single API query returned too many records. By using `take`, you can limit the number of records returned by each query made by the function. The function will automatically make as many queries as it needs to retrieve the data in chunks not exceeding the size determined by `take`. The default value and maximum number of records that the API will ever return for a single query is `10000` so if you do get a 500 error, try setting `take` to less than `10000`.

```{r, eval=FALSE}
# Setting the take to 5000 to be safe, but the API should respond just fine if the
# take were unspecified here because it will only ever return 10000 records per query.
# NOTE: It is absolutely not recommended that you attempt to download the entire
# data set this way. Please contact the data managers at the LDC if you need a full
# data set.
lpi_data <- fetch_ldc(data_type = "lpi",
                      take = 5000,
                      verbose = TRUE)
```

#### exact_match (deprecated)

Prior to v2.0.0, `fetch_ldc()` supported partial string matching in queries. While this is no longer the case, the functionality can be approximated using two separate calls to `fetch_ldc()`. For example, if you want to retrieve all the indicator values from sampling events that took place in a Major Land Resource Area (MLRA) there's no MLRA variable to query by but every ecological site ID contains the MLRA that the ecological site occurs in. The first `fetch_ldc()` call would request an entire table with the ecological site IDs in it and then the second would request the target data for PrimaryKeys where the MLRA ID showed up in the associated ecological site IDs.

```{r, eval=FALSE}
# Retrieving indicator values for all sampling events that took place in
# the MLRA 036X (southwestern Colorado, northern New Mexico, and southeastern Utah)
# based on the EcologicalSiteID. This could also be done using a polygon of the MLRA
# boundary with fetch_ldc_spatial().

# This will return all the header records for Colorado, New Mexico, and Utah
relevant_states <- c("CO", "NM", "UT")
current_query_parameters <- list("State" = list("in" = relevant_states))
current_headers <- fetch_ldc(data_type = "headers",
                             query_parameters = current_query_parameters)

# This will filter those to only records where the EcologicalSiteID value contains
# the MLRA ID and then keep only the PrimaryKeys associated with those plots.
mlra_primarykeys <- dplyr::filter(.data = current_headers,
                                  stringr::str_detect(string = EcologicalSiteID,
                                                      pattern = "036X")) |>
  dplyr::pull(.data = _,
              var = "PrimaryKey")

# The PrimaryKeys for plots in the MLRA can then be used to get the indicator table
# records
mlra_036x_indicators <- fetch_ldc(data_type = "indicators",
                                  query_parameters = list("PrimaryKey" = list("in" = mlra_primarykeys)))
```

## fetch_ldc_spatial()

The LDC API does not currently directly support spatial queries, but `fetch_ldc_spatial()` makes it possible to retrieve records which fall within user-supplied polygons. This is a wrapper function for `fetch_ldc()` that simply automates a few steps to streamline spatial querying. It does this by converting the header table from the LDC into a POINT sf object and returning all requested data that correspond to the `PrimaryKey` values associated with points that intersect the provided polygons.

The two required arguments are `polygons` and `data_type`. The polygons must be an sf POLYGON or MULTIPOLYGON object (see the [package sf](https://cran.r-project.org/web/packages/sf/index.html) for additional information) and `data_type` must correspond to one of the tables in the LDC (see `fetch_ldc()`). There is no need to combine or dissolve the polygons or to specify an attribute or variable associated with them: the function will return all points that intersect any polygon in the object.

Like `fetch_ldc()`, this function also accepts the arguments `timeout`, `take`, and `verbose` (see above). In particular, `take` may be important to specify or adjust because the function is first identifying relevant `PrimaryKey` values and then using those for followup queries to retrieve the requested data. This means that, depending on the scale of the polygons and the density of data, there may tens of thousands of records to return. If the server is returning 500 errors, start by lowering the value for `take`.

```{r, eval=FALSE}
# Retrieving indicator values for all sampling events that took place in
# vegetation removal treatments of interest
# Read in the polygons as an sf object
treatment_polygons <- sf::st_read(dsn = "vegetation_treatments.gdb",
                                  layer = "removal_treatment_polygons")
# Use the sf object to retrieve only relevant data
treatment_indicator_data <- fetch_ldc_spatial(polygons = treatment_polygons,
                                              data_type = "indicators",
                                              verbose = TRUE)

# Retrieving line-point intercept data for all sampling events that took place
# in vegetation removal treatments
treatment_lpi_data <- fetch_ldc_spatial(polygons = treatment_polygons,
                                        data_type = "lpi",
                                        verbose = TRUE)
```

## fetch_ldc_ecosite()

One of the most common attributes used to request data is the associated ecological site ID. This function is a wrapper for `fetch_ldc()` that only queries by one or more ecological site IDs for the sake of simplifying those queries.

The minimum required arguments are just `ecosite_ids` which are compared against the values in `EcologicalSiteID` in the headers table and `data_type` which specified which table to retrieve associated records from. Additional supported arguments for how the queries are submitted can be found in the documentation.

```{r, eval=FALSE}
# Retrieving gap values for all sampling events that took place in
# the ecological sites R036XB006NM and R036XB007NM
ecosite_gap_data <- fetch_ldc_ecosite(data_type = "gap",
                                      ecosite_ids = c("R036XB006NM", "R036XB007NM"),
                                      verbose = TRUE)

# Retrieving soil stability values for all sampling events that took place in
# the MLRA 036X (southwestern Colorado, northern New Mexico, and southeastern Utah)
# This will return all records where the value in EcologicalSiteID contains "036X"
mlra_gap_data <- fetch_ldc_ecosite(keys = "036X",
                                   data_type = "soilstability",
                                   exact_match = FALSE,
                                   verbose = TRUE)
```
